\speculake: Hiding Malware in Speculative Execution

### 2018-07-26 paper/related-work.tex
trim related work section

The term \textit{weird machine} comes from work done by Bratus \textit{et al.}
composing bugs and features of a system to create arbitrary computation with
interesting properties~\cite{weird_machines,bratus2011exploit,weird_exploits}.
This encompasses many traditional computer vulnerabilities including buffer
overflows~\cite{buffer_overflow}, format string
exploits~\cite{format_string_exploit}, and return oriented
programming~\cite{shacham2007geometry}. These exploits demonstrate performing
arbitrary computation in places that were not originally intended to act as
computational entities.

As another example, Bangert~\textit{et al.} demonstrate that a Turing complete
weird machine can be constructed using the page fault handling mechanism in
Intel's IA32 architecture~\cite{bangert2013page}.

Unexpected functionality like this is abundant in modern
processors~\cite{d2015exploiting} as multiple threads share system resources
cooperatively and optimizations are often done across the process
isolation boundary. \speculake takes advantage of cooperative hardware
resources, and constructs a hardware weird machine
using speculative branching and \texttt{Flush+Reload}.

\smallskip

Hardware minutia have also been used to identify and fingerprint execution
environments --- in malware this is know as a red-pill~\cite{red-pill} and 
is typically used by malware to detect when it is under emulation or
a debugging environment~\cite{lindorfer2011detecting, balzarotti2010efficient}. Paleari
\textit{et al.} automate the process of discovering red-pills by
fuzzing emulated and physical environments and identifying
differences between the two~\cite{paleari2009fistful}.
%This is still by no means an exhaustive list of red-pills for a given 
%processor as many rely on timing or race conditions in physical 
%environments.

@@

\smallskip

@@

The \speculake model relies critically on cache side channel attacks to 
communicate information from the speculative world. Cache side-channels have 
been used against a number of processors at essentially all different 
caching levels -- L1 data cache~\cite{percival2005cache,zhang2012cross,osvik2006cache}, 
L1 instruction cache~\cite{aciiccmez2010new}, 
L2 cache~\cite{ristenpart2009hey,percival2005cache}, 
LLC~\cite{ristenpart2009hey,liu2015last}, 
and branch prediction cache~\cite{aciiccmez2007power}. 
These attacks have also been performed in 
cloud~\cite{ristenpart2009hey,zhang2012cross}, 
browser~\cite{oren2015spy,google_cache_browser},
and security critical encryption~\cite{yarom2014recovering,tromer2010efficient}
environments, demonstrating how ubiquitous hardware side-channel exploits can be.

@@

% Discovery & Followup
Given the numerous directions that were being investigated in reverse
engineering Intel processors, multiple teams discovered the Spectre
vulnerabilities concurrently including the teams associated with Kocher
\textit{et al.}~\cite{spectre}, the Project Zero team at
Google~\cite{project_zero}, and various
others~\cite{evtyushkin2018branchscope,maisuradze2018speculose}. 
%These works propose multiple variants making use of branch prediction.
Follow 
up work on speculative execution side channels has also demonstrated
web-based vulnerabilites~\cite{genkin2018drive} as well as hardware isolation  
leaks in Intel's SGX~\cite{spectre_sgx}. At the time of writing multiple sources
have claimed new speculative execution related exploits termed \textit{Spectre-NG}
have been reported to Intel and assigned CVE numbers, though details are
not yet available~\cite{spectre_ng_gossip}.
%We have no details
%on content, though Intel has labeled four as critical and another four 
%as intermediate security flaws~\cite{spectre_ng_gossip}.

@@

%Much like cache side channel vulnerabilities speculative execution is relatively
%critical to processor performance and requires significant software and hardware 
%updates to fully address. 
Multiple measures have been proposed to combat the 
individual variants of Spectre.

One popular approach uses compiler level software patches to wrap sensitive 
function calls. Retpoline replaces indirect jumps with \texttt{retq}
instructions that read from the cached stack (and are thus less likely to be
mis-speculated)~\cite{retpoline}.
Gruss~\textit{et al.} propose the KAISER software patch to enforce address space
isolation for kernel and user space proceses~\cite{gruss2017kaslr}. 
Other proposals have considered hypervisor enforced time domain separation at varying 
granularity~\cite{renau2018securing}, and fuzzy clocks that would reduce the accuracy 
of any time based exfiltration pathway~\cite{hu1992reducing}. However, none of these defenses
fundamentally defeat the use of a branch predictor as a cooperative mechanism
spanning multiple processes. Intel has released microarchitecture updates for
its most recent CPUs, but older architectures do not support these updates and
will remain vulnerable~\cite{intel-micro-patch}.

Trippel~\textit{et al.} proposes a method of faithfully simulating hardware
bugs, given a formal model specifying the
microarchitecture~\cite{trippel2018meltdownprime}. However, such models are
unlikely to be released by CPU manufacturers, as they generally involve
closely-guarded trade secrets, making this a less accessible solution for the
research community.

>>>>

\textit{Weird machines} are machines composing bugs and features of a system to
create arbitrary computation with interesting
properties~\cite{weird_machines,bratus2011exploit,weird_exploits}. This
encompasses many traditional computer vulnerabilities~\cite{buffer_overflow,
format_string_exploit, shacham2007geometry}. These exploits demonstrate
performing arbitrary computation in places that were not originally intended to
act as computational entities.

Unexpected functionality such as~\cite{bangert2013page} is abundant in modern
processors~\cite{d2015exploiting} as multiple threads share system resources
cooperatively and optimizations are often done across the process
isolation boundary. \speculake takes advantage of cooperative hardware
resources, and constructs a hardware weird machine
using speculative branching and \texttt{Flush+Reload}.

Hardware minutia have also been used to identify and fingerprint execution
environments---in malware this is know as a red-pill~\cite{red-pill} and is
typically used by malware to detect when it is under emulation or a debugging
environment~\cite{lindorfer2011detecting, balzarotti2010efficient,
paleari2009fistful}

@@

@@

Cache side-channels have been used against a number of processors at essentially
all different caching levels---L1 data
cache~\cite{percival2005cache,zhang2012cross,osvik2006cache}, L1 instruction
cache~\cite{aciiccmez2010new}, L2
cache~\cite{ristenpart2009hey,percival2005cache},
LLC~\cite{ristenpart2009hey,liu2015last}, and branch prediction
cache~\cite{aciiccmez2007power}. These attacks have also been performed in the
cloud~\cite{ristenpart2009hey,zhang2012cross},
browser~\cite{oren2015spy,google_cache_browser}, and security critical
encryption~\cite{yarom2014recovering,tromer2010efficient} environments,
demonstrating how ubiquitous hardware side-channel exploits can be.

@@

% Discovery & Followup
Given the numerous directions that were being investigated in reverse
engineering Intel processors, multiple teams discovered the Spectre
vulnerabilities concurrently~\cite{spectre, project_zero, evtyushkin2018branchscope, maisuradze2018speculose}.
%These works propose multiple variants making use of branch prediction.
Follow 
up work on speculative execution side channels has also demonstrated
web-based vulnerabilites~\cite{genkin2018drive} as well as hardware isolation  
leaks in Intel's SGX~\cite{spectre_sgx}. At the time of writing multiple sources
have claimed new speculative execution related exploits termed \textit{Spectre-NG}
have been reported to Intel and assigned CVE numbers, though details are
not yet available~\cite{spectre_ng_gossip}.
%We have no details
%on content, though Intel has labeled four as critical and another four 
%as intermediate security flaws~\cite{spectre_ng_gossip}.

@@

%Much like cache side channel vulnerabilities speculative execution is relatively
%critical to processor performance and requires significant software and hardware 
%updates to fully address. 
% Multiple measures have been proposed to combat the 
% individual variants of Spectre.

% One popular approach uses compiler level software patches to wrap sensitive 
% function calls. Retpoline replaces indirect jumps with \texttt{retq}
% instructions that read from the cached stack (and are thus less likely to be
% mis-speculated)~\cite{retpoline}.
% Gruss~\textit{et al.} propose the KAISER software patch to enforce address space
% isolation for kernel and user space proceses~\cite{gruss2017kaslr}. 
% Other proposals have considered hypervisor enforced time domain separation at varying 
% granularity~\cite{renau2018securing}, and fuzzy clocks that would reduce the accuracy 
% of any time based exfiltration pathway~\cite{hu1992reducing}. However, none of these defenses
% fundamentally defeat the use of a branch predictor as a cooperative mechanism
% spanning multiple processes. Intel has released microarchitecture updates for
% its most recent CPUs, but older architectures do not support these updates and
% will remain vulnerable~\cite{intel-micro-patch}.

% Trippel~\textit{et al.} proposes a method of faithfully simulating hardware
% bugs, given a formal model specifying the
% microarchitecture~\cite{trippel2018meltdownprime}. However, such models are
% unlikely to be released by CPU manufacturers, as they generally involve
% closely-guarded trade secrets, making this a less accessible solution for the
% research community.
